---
title: "Day1_answers"
author: "Alex McFarland"
date: "8/11/2021"
output: html_document
---

```{r setup, include=FALSE}

library(ggtree)
library(ggplot2)
library(dplyr)
library(treeio)
library(phytools)
library(ape)
library(ggpubr)

setwd('/Users/owlex/Dropbox/Documents/Northwestern/rcs_consult/r_phylogenetics_workshop/r_phylogenetics_worshop')


```

# Day 1 answers

---
### Exercise 1

1.  Use `ape::read.tree()` to read in the file `raw_enoyl_seqs.nwk` and assign it to the variable `my_nwk`

```{r}

my_nwk <- ape::read.tree('raw_enoyl_seqs.nwk')

```


2. Display the `my_nwk` tree using `ggtree()`

```{r}

ggtree(my_nwk)

```

3. Show the `geom_tiplab()` for each leaf in the `ggtree()` visualization of `my_nwk`

```{r}

ggtree(my_nwk)+
  geom_tiplab()

```


4. Modify the code from part 3 so that the x-axis is rescaled at `x=7` using `geom_treescale()`

```{r}

ggtree(my_nwk)+
  geom_tiplab()+ # show tip labels
  geom_treescale(x=7) # recenter the tree
  
```

---

### Exercise 2 

1. Use `ape::read.tree()` to read in the tree in `'RAxML_bipartitions.raw_enoyl_seqs'` and store it to the variable `my_tree_raxml`.

```{r}

my_tree_raxml <- ape::read.tree('RAxML_bipartitions.raw_enoyl_seqs')

```


2.  Use `ape::root()` to root the tree to the outgroup `sp|P0AEK4|FABI_ECOLI` and store it in the variable `my_rooted_tree`

```{r}

my_rooted_tree <- ape::root(my_tree_raxml ,outgroup='sp|P0AEK4|FABI_ECOLI')

```

3. Visualize the tree using `ggtree()` with with annotation layers `geom_treescale(x=10)`, `geom_tiplab()`, and `geom_nodelab(aes(label=label))`

```{r}

ggtree(my_rooted_tree)+
  geom_tiplab()+
  geom_nodelab(aes(label=label))+
  geom_treescale(x=10)

```

### Exercise 3

1. Read in two trees: `'raw_enoyl_seqs.nwk'` and `'RAxML_bipartitions.raw_enoyl_seqs'` with `ape::read.tree()`. Store in the variables `tree1_fasttree` and `tree2_raxml`

```{r}

tree1_fasttree <- ape::read.tree('raw_enoyl_seqs.nwk')

tree2_raxml <- ape::read.tree('RAxML_bipartitions.raw_enoyl_seqs')

```

2. For both trees, use `ape::root()` to root the tree to the leaf `'tr|E6KYQ2|E6KYQ2_9PAST'`

```{r}

tree1_fasttree <- ape::root(tree1_fasttree ,outgroup='tr|E6KYQ2|E6KYQ2_9PAST')

tree2_raxml <- ape::root(tree2_raxml ,outgroup='tr|E6KYQ2|E6KYQ2_9PAST')

```

3. Run `ggtree()` on both sets of trees and store them in variables `tree1_fasttree_visual` and `tree2_raxml_visual`

```{r}

tree1_fasttree_visual <- ggtree(tree1_fasttree)+
  geom_tiplab(size=2,align = TRUE,offset=.3)+
  geom_tippoint(aes(color=label))+
  geom_treescale(x=7)+
  theme(legend.position='none')

tree2_raxml_visual <- ggtree(tree2_raxml)+
  geom_tiplab(size=2,align = TRUE,offset=.3)+
  geom_tippoint(aes(color=label))+
  geom_treescale(x=7)+
  theme(legend.position='none')

```

4. Use `ggpubr::ggarrange()` to plot both trees side by side

```{r}

ggpubr::ggarrange(tree1_fasttree_visual, tree2_raxml_visual)

```

5. Add the annotation layer `scale_x_reverse()` to flip the orientation of the tree visualization

```{r}

tree2_raxml_visual <- ggtree(tree2_raxml)+
  geom_tiplab(size=2,align = TRUE,offset=-.1,hjust=1)+
  geom_tippoint(aes(color=label))+
  geom_treescale(x=-7)+
  theme(legend.position='none')+
  scale_x_reverse()

tree2_raxml_visual

```

6. Use `ggpubr::ggarrange()` to plot the flipped `tree2_raxml_visual` next to `tree1_fasttree_visual`

```{r}

ggpubr::ggarrange(tree1_fasttree_visual, tree2_raxml_visual)

```


----

### Exercise 4

1. Read in the `'RAxML_bipartitions.raw_enoyl_seqs'` tree using `ape::read.tree()`. Store in the variable `tree1_raxml`. Root the tree to `'tr|E6KYQ2|E6KYQ2_9PAST'` using `ape::root`. 
Make a tree showing each internal node label using `geom_nodelab()`

```{r}

tree1_raxml <- ape::read.tree('RAxML_bipartitions.raw_enoyl_seqs')
tree1_raxml <- ape::root(tree1_raxml ,outgroup='tr|E6KYQ2|E6KYQ2_9PAST')

tree1_raxml_visual <- ggtree(tree1_raxml)+
  geom_nodelab(aes(label=node),nudge_x=-.15,nudge_y=.4,size=2,color='blue')+
  geom_tiplab(size=3)+
  geom_treescale(x=7)

tree1_raxml_visual

```


2. Zoom in on node 35 and store the zoom in visual in the variable `zoomin_visual1`

```{r}

zoomin_visual1 <- viewClade(tree1_raxml_visual, node=36)

zoomin_visual1

```

---

