---
title: "day3_answers"
author: "Alex McFarland"
date: "8/13/2021"
output: html_document
---


```{r setup}

#install.packages('ggtree')
#install.packages('ggplot2')
#install.packages('dplyr')
#install.packages('treeio')
#install.packages('phytools')
#install.packages('ape')
#install.packages('ggpubr')
#install.packages('ggnewscale')

library(ggtree)
library(ggplot2)
library(dplyr)
library(treeio)
library(phytools)
library(ape)
library(ggpubr)
library(ggnewscale)
library(ggstance)

setwd('/Users/owlex/Dropbox/Documents/Northwestern/rcs_consult/r_phylogenetics_workshop/r_phylogenetics_worshop') # change this to your R Markdown's file path

```

---

### Exercise 1

Edit the above visualization so that the node pies are a different color, gene names are used, and a zoomed in visualization is provided.

1. Associate `df_meta` with `base_tree` using the `%<+%` operator. Set the `geom_tiplab()` `aes(label=gene_name)`

```{r}

tree_vis_ex <- base_tree%<+%df_meta+
  geom_inset(pies,width=.1, height=.1)+
  geom_tiplab(size=2.6,align = TRUE,offset=.05, aes(label=gene_name))+
  geom_treescale(x=3, color=NA)

tree_vis_ex

```

2. Use the `lapply()` function to changethe color of the pie chart to `'orange'` and `'blue'`. 

Add the new `pies2` to `geom_inset()`

```{r}

pies <- nodepie(df_ancstats,cols=c('high','low'),alpha=0.8)
pies2 <- lapply(pies, function(g) g + scale_fill_manual(values = c('orange','blue')))

tree_vis_ex2 <- base_tree%<+%df_meta+
  geom_inset(pies2,width=.1, height=.1)+
  geom_tiplab(size=2.6,align = TRUE,offset=.05, aes(label=gene_name))+
  geom_treescale(x=3, color=NA)

tree_vis_ex2

```

3. Use `viewClade()` to zoom in on `node=25` and store in `zoom_in_vis`. After, use `ggpubr::ggarrange()` view `tree_vis_ex2` and `zoom_in_vis` side by side.

```{r}

zoom_in_vis <- viewClade(tree_vis_ex2, node=25)

ggpubr::ggarrange(tree_vis_ex2, zoom_in_vis)

```